---
title: "Intro to R"
author: "Jeffrey Verbeurgt"
date: "28 March 2019"
output:
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## 1. Getting Started

- What is R?
-	Why use R?
- Why use Rstudio?
- Basic R syntax
- Write your first script

## What is the R-project?

- A language and environment for statistical computing and graphics
- A community of programmers, users and academics
- A powerful tool that makes publication-quality graphics and data visualizations
- A collection of freely distributed add-on packages
- A toolbox for extensive automation of your work


<div class="notes">
For Python users: the well-known R ggplot2-package is an excellent alternative for matplotlib or plotnine

The amount of add-on packages is huge, if you want to do something "special": there probably exist a package for it

One uses R for automation of your work, saving you time to do real work

</div>


## Why use R?

- Automation of analyses
- Sharing analyses & scripts
- Reproducible analyses
- State of the art packages
- Quick exploration of data

<div class="notes">
Don't use R if you expect point-and-click ease like SPSS

Don't spend time on the steep learning curve if you intend to do only one single analysis

The amount of state-of-the-art packages is unbelievable, however don't use models you don't understand
</div>

## Why use Rstudio?

- Gives a good overview with the 4 work areas
    - Source
    - Console
    - Environment (+ history)
    - Files, Plots, Packages, Help and Viewer
- Easier to use than R console
- Create Notebook, Markdown (slides, pdf), Shiny Web App,...

<div class="notes">
In 'source' you can create scripts etc
In 'console': type in 1 + 1

R Markdown Notebook: same as the well known Python Jupyter Notebook
Markdown: used for these slides, also easy to create reports / course notes /...
Shiny: Filipe will show more of this later today
</div>

## Basic R syntax
Let's start with some basic commands in the console:
```{r echo=TRUE, eval = FALSE}
print("Hello world!")
```
R responds immediately
```{r echo=FALSE, eval = TRUE}
print("Hello world!")
```

Now let's try to calculate the sum of five numbers:
```{r echo=TRUE, eval = TRUE}
1 + 2 + 3 + 4 + 5
```

## Basic R syntax
You took the sum of 5 numbers, in a vector format this is:
```{r echo=TRUE, eval = TRUE}
c(1, 2, 3, 4, 5)
```

The sum of these 5 numbers is a lot of typing:
```{r echo=TRUE, eval = TRUE}
c(1:5)
```
And in R it's easy to take the sum of a vector:
```{r echo=TRUE, eval = TRUE}
sum(1:5)
```

## Write your first script
Choose 'File' -> 'New File' -> 'R Script' 
```{r echo=TRUE, eval = TRUE}
x <- 1:5
x
```
```{r echo=TRUE, eval = TRUE}
y <- 10
x + y
```


## Write your first script
Note: calculations only get stored when they are assigned
```{r echo=TRUE, eval = TRUE}
x
```
```{r echo=TRUE, eval = TRUE}
y
```
```{r echo=TRUE, eval = TRUE}
z <- x + y
z
```

## Write your first script
Now let's make a new script
```{r echo=TRUE, eval = FALSE}
h <- "Hello,"
yourName <- readline("What is your name?")
print(paste(h, yourName))
```

```{r echo=FALSE, eval = TRUE}
h <- "Hello,"
yourName <- "Jeffrey"
```

```{r echo=FALSE, eval = TRUE}
print(paste(h, yourName))
```


## 2. First programming with R

- R is vector-based
- R functions
- R packages
- Basic Arithmetic, Indexing, Comparing values
- Code structure and conventions


## R is vector-based
R is clever when it comes to vectors:
```{r echo=TRUE, eval = TRUE}
title <- c("King", "Queen")
firstname <- c("Philippe", "Mathilde")
paste(title, firstname)
```

```{r echo=TRUE, eval = TRUE}
title <- c("King")
firstname <- c("Philippe", "Mathilde")
paste(title, firstname)
```

## functions in R
- A lot of functions in R
- Always followed by parentheses ()
- They often have arguments placed between ()
```{r echo=TRUE, eval = FALSE}
print()
```
- Check out the Help page
```{r echo=TRUE, eval = FALSE}
?print()
```
```{r echo=TRUE, eval = TRUE}
print("Hello world!")
```


## Packages in R
- The big power of R: packages
- Packages are groups of functions written by other people
- Packages can be downloaded from repositories
```{r echo=TRUE, eval = FALSE}
install.packages("stringr")
```
```{r echo=TRUE, eval = TRUE}
library("stringr")
```
```{r echo=TRUE, eval = FALSE}
?stringr
```

<div class="notes">
Repositories are special websites containing packages

Let me show you an example of a package: stringr
https://cran.r-project.org/web/packages/stringr/index.html
</div>

## Indexing
```{r echo=TRUE, eval = TRUE}
stringList <- c("Geography", "Rules", "!")
paste(stringList[1], stringList[3])
stringList[-2]
integerList <- c(1, 5, 10)
integerList[1] + integerList[3]
```

## Comparing values
```{r echo=TRUE, eval = TRUE}
1 < 5
1 >= 5
"Geography" != "Engineering"
"Geography" > "Engineering"
```


## Comparing values
```{r echo=TRUE, eval = TRUE}
string1 <- "Geography"
string2 <- "geography"
identical(string1, string2)
```


## Code structure and conventions
- Choose a correct name
    - names must start with a letter or a dot
    - names should only contain letters, numbers, underscore, characters and dots
    - you can't use special keywords like break, else, FALSE,...
    - R is case sensitive
    - Choose a clear name that you understand (even a year later)
    - Choose a naming style: camelCase, dotted style,..

## Code structure and conventions
- Structuring your code
    - Use a quotation mark to start a string
    ```{r echo=TRUE, eval = FALSE}
"This is the first line
    This is the second line"
```
    - You can end an incomplete line with an operator
    ```{r echo=TRUE, eval = TRUE}
z <- 25 + 5 +
      10 + 2
z
```    

## Code structure and conventions
    - You can open a parenthesis for a function
```{r echo=TRUE, eval = TRUE}
paste("Hello",
      "world!")
```   
    - Use comments for extra information
```{r echo=TRUE, eval = TRUE}
paste("Hello",
      "world!")
```


## 3. More Programming with R

- Naming values in a vector
- R objects (factors, matrix, data frame, list, tibble)
- If-else and for-loops
- One big apply-family


## Naming values in a vector
```{r echo=TRUE, eval = TRUE}
month.days <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31)
month.name #is a built-in dataset
names(month.days) <- month.name
month.days
```

## R objects: factors
- Factors are categorical vectors with only a few levels
    - hot/cold, male/female, small/medium/large
```{r echo=TRUE, eval = TRUE}
status <- c("Lo", "Hi", "Med", "Med", "Hi")
factor(status)
ordered.status <- factor(status, levels = c("Lo", "Med", "Hi"), 
                         ordered = TRUE)
ordered.status
```

<div class="notes">
factors: hot/cold, male/female, small/medium/large
</div>

## R objects: list
think of a list as a big box filled with a set of bags containing all kinds of different stuff
```{r echo=TRUE, eval = TRUE}
exampleList <- list(c(1:5), c("A", "B", "C", "D"), c(FALSE, FALSE))
exampleList
```

## R objects: matrices
A matrix can only contain one datatype
```{r echo=TRUE, eval = TRUE}
exampleMatrix <- matrix(c("A", 1, TRUE, 2), ncol = 2, nrow = 2)
exampleMatrix
typeof(exampleMatrix[1,1])
typeof(exampleMatrix[1,2])
```


## R objects: dataframes
- Generalized version of a matrix
- can contain heterogeneous data
- is a list of vectors of equal length
```{r echo=TRUE, eval = TRUE}
#iris is a built-in dataset
#head() is used to show the first 6 lines
head(iris)
```
## R objects: dataframes
- Generalized version of a matrix
- can contain heterogeneous data
- is a list of vectors of equal length
```{r echo=TRUE, eval = TRUE}
#iris is a built-in dataset
#str() is used to show the structure
str(iris)
```

## If-else
```{r echo=TRUE, eval = TRUE}
x <- 5
if(x < 10){
  print("x is smaller than 10")
} else{
  print("x is not smaller than 10")
}
  
```
```{r echo=TRUE, eval = TRUE}
x <- 20
if(x < 10){
  print("x is smaller than 10")
} else{
  print("x is not smaller than 10")
}
  
```

## for-loops
```{r echo=TRUE, eval = TRUE}
for(month in month.name){
  print(month)
}
```

## One big apply-family
- apply(), sapply() and lapply() help you in applying a function over columns or rows
- first argument is always the the matrix/array, data frame or list
- only for apply() you specify in which dimension the function should work (1=row, 2=column)
- sapply() and lapply() work only on data frames or lists
- check out the iris dataset
```{r echo=TRUE, eval = TRUE}
head(iris)
```

## One big apply-family
```{r echo=TRUE, eval = TRUE}
apply(iris, 2, min)
```

```{r echo=TRUE, eval = TRUE}
sapply(iris, class)
```

## One big apply-family

```{r echo=TRUE, eval = TRUE}
lapply(iris, class)
```















